Missing the Point\(\colon\) Non-Convergence in Multiple Imputation Procedures

### 2020-04-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Before removing 9 of the 12 plots in the manuscript with the different missingness percentages.

With multiple imputation, missing data are 'imputed' (i.e., filled in) several times by means of an iterative algorithmic procedure. The method ultimately results in an aggregated statistical inference. For the theoretical foundation of MI, see @rubin87. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. @buur18.

based on a model of the observed databy means of an iterative algorithm 
is the use of MICE: multiple imputation by chained equations. MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations). Imputed values are obtained by modeling the (distribution of the) missing data conditional on the observed data. The technique is therefore also called 'fully conditional specification' (FCS).
<!-- several times, and inferences are aggregated across imputed datasets.  -->
<!-- and statistical inference is performed on each imputed dataset. Variation between the results refelect uncertainty due to missingness  -->
<!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification). FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".  -->

@@

There is no scientific consensus on the convergence properties of FCS algorithms [@taka17]. Some default techniques in `mice` might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

Convergence cannot be determined for sure, because MICE is an MCMC method, and convergence of MCMC is not from scalar to a point, but from one distribution to another. Generated values will vary even after convergence. Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96].

Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, $\S$ 6.5.2]. Visually inspecting imputation chains (**for signs of non-mixing or non-stationarity**) can only diagnose severely pathological cases of non-convergence. Therefore diagnostic evaluation of convergence would be preferred. However, not known whether MCMC diagnostics are appropriate for MICE ... The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. How can non-convergence be diagnosed? How many iterations are sufficient/required? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? Defaults in: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

@@

The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine (**the chain length required to cancel out trending**) the effect of trending within chains.

@@

Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). @buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03]. This method is outside of the scope of the current study (**not anymore!**).

@@

@@

>>>>

The multiple imputation method can be roughly divided into 3 steps. First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times. Next, the analysis of scientific interest is performed on each of the completed datasets. And finally, the results of these analyses are pooled into a single aggregated statistic. The theoretical foundation of these streps was laid by Rubin [1972, -@rubin87]. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. @buur18. In this paper, we focus on the first of these steps: the algorithmic procedure with which the missing values are imputed.

<!-- based on a model of the observed databy means of an iterative algorithm  -->
<!-- is the use of MICE: multiple imputation by chained equations. MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations). Imputed values are obtained by modeling the (distribution of the) missing data conditional on the observed data. The technique is therefore also called 'fully conditional specification' (FCS). -->
<!-- several times, and inferences are aggregated across imputed datasets.  -->
<!-- and statistical inference is performed on each imputed dataset. Variation between the results refelect uncertainty due to missingness  -->
<!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification). FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".  -->

@@

There is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default techniques might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully. The current practice is visually inspecting imputation chains for signs of non-convergence, but this may be challenging to the untrained eye [@buur18, $\S$ 6.5.2]. Moreover, visually inspecting imputation chains can only diagnose severely pathological cases of non-convergence. Therefore, diagnostic evaluation of convergence would be preferred.

Convergence cannot be determined for sure, because MICE (**introduce MICE before!**) is a type of MCMC method, and convergence of MCMC algorithms is not from a scalar to a point, but from one distribution to another. Therefore, generated values will vary even after convergence. The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose severe non-convergence.
<!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

Several MCMC convergence diagnostics exist, but it is not known whetherthese are appropriate for MICE. The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. The open questions are, e.g.: How can non-convergence be diagnosed? How many iterations are sufficient/required? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

@@

The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine  the effect of trending within chains (**or the chain length required to cancel out trending?**).

@@

Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). @buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03]. This method is outside of the scope of the current study (**not anymore?**).

@@

## Univariate estimates and convergence diagnostics

The bias in the estimates of the variable means show little to no difference between simulation conditions. It doesn't seem to matter how many iterations you use in the mice algorithm, the estimates are unbiased. Similarly, the bias in the estimated variances is more or less stable across simulation conditions. These univariate quantities appear to be unaffected by the number of iterations.

When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions. Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01). The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$. (With the default number of iterations (maxit = 5), this dip in Rhat values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the Rhat value is around $T=30$.
<!-- Rhat values are indistinguishable between simulation conditions when $T>30$.  -->
<!-- The R hat values seemed to flatten off between 20 and 30 iterations.  -->

Auto-correlations indicate no sign of trending within imputation chains. In most simulation conditions AC is smaller than or about equal to zero. Across simulation conditions, however, the autocorrelation curve does not trend towards 0. Autocorrelation values plateau off at a value of around .1 . This is a small positive autocorrelation, which would indicate some trending within chains.

The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends and are therefore not discussed separately. Taken together,  univariate estimates seem robust across simulation conditions. There is no clear effect of the number of iterations on the bias in these estimates, while the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).

## Multivariate estimates

There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four. In simulation conditions where $T > 5$ there is little to no bias in the estimated regression coefficients. In most simulation conditions nominal coverage is obtained (i.e., coverage rates of .95) for the confidence intervals around the regression coefficients. Conditions with only one or two iterations show some undercoverage. Since confidence interval width is stable across conditions, the undercoverage may be attributed to the bias in the estimated regression coefficients.

If we look at the estimated proportion of explained variance in outcome variable Y we see that the coefficient of determination is underestimate estimated in conditions where the number of iterations is equal to two or less, and slightly overestimated in conditions where the number of iterations is equal to three or more.

In short, we see that the minimum number of iterations required to obtain unbiased, confidence valid regression estimates is 5. This value, however, is dependent on the percentage of missing values. E.g., with 95% of cases having missing data we need at least seven iterations to obtain unbiased results.

<!-- Bias in the estimated $R^2$ values does not improve  -->

@@

<!-- For 25% miss, no bias in estimated means, Rh flat from 30 it. (but <1.1 for it>11), AC never high but stable after 40 it. Bias in estimated variances only if it=1 or maybe 2?, while Rh of chain variances is flat around it 25 (smaller than 1.1 for it>9), and AC never positive, flat around it 30. Bias in estimated reg. coeff. only pronounced for the first iteration., while coverage rates are ok immediately on average, variance explained is marginally underestimated (bias of 0.5% on an effect of 19.25%). -->

```{r include=FALSE}
# chain means
dat[-1, 6:9] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2

# chain variances
dat[-1, 26:29] %>% mutate(max_Rh = apply(., 1, max)) %>% .$max_Rh<1.2

# Results [OLD, JUST FOR REFERENCE!]

### 2020-04-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Before removing 9 of the 12 plots in the manuscript with the different missingness percentages.

With multiple imputation, missing data are `imputed' (i.e., filled in)
several times by means of an iterative algorithmic procedure. The method
ultimately results in an aggregated statistical inference. For the
theoretical foundation of MI, see \citet{rubin87}. For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~\citet{buur18}.

based on a model of the observed databy means of an iterative algorithm
is the use of MICE: multiple imputation by chained equations. MICE is a
type of multiple imputation procedure--a framework proposed by
\citet{rubin87}--in which missing datapoints are `imputed' (i.e., filled
in) by means of an iterative algorithm (chained equations). Imputed
values are obtained by modeling the (distribution of the) missing data
conditional on the observed data. The technique is therefore also called
`fully conditional specification' (FCS).

There is no scientific consensus on the convergence properties of FCS
algorithms \citep{taka17}. Some default techniques in \texttt{mice}
might not yield converged states at all \citep{murr18}. Therefore,
algorithmic convergence should be monitored carefully.

Convergence cannot be determined for sure, because MICE is an MCMC
method, and convergence of MCMC is not from scalar to a point, but from
one distribution to another. Generated values will vary even after
convergence. Therefore we can only diagnose non-convergence, not
convergence \citep{hoff09}. ``a weak diagnostics is better than no
diagnostic at all'' \citep{cowl96}.

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye
\citep[\(\S\) 6.5.2]{buur18}. Visually inspecting imputation chains
(\textbf{for signs of non-mixing or non-stationarity}) can only diagnose
severely pathological cases of non-convergence. Therefore diagnostic
evaluation of convergence would be preferred. However, not known whether
MCMC diagnostics are appropriate for MICE \ldots{} The application of
convergence diagnostics to MI methods has not been systematically
studied \citep{buur18}. How can non-convergence be diagnosed? How many
iterations are sufficient/required? What are the effects of
non-convergence on estimates, predictions and inference? What is an
appropriate non-convergence diagnostic? Are the default number of
iterations sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 10
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

@@

The stationarity component of convergence may be evaluated with
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
\citeyearpar{raft91} procedure to determine (\textbf{the chain length
required to cancel out trending}) the effect of trending within chains.

@@

Note that all of these methods evaluate the convergence of univariate
scalar summaries (e.g., chain means or variances). These convergence
diagnostics cannot diagnose convergence of multivariable statistics
(i.e., relations between scalar summaries). \citet{buur18} proposed to
implement multivariable evaluation through eigenvalue decomposition
\citep{mack03}. This method is outside of the scope of the current study
(\textbf{not anymore!}).

@@

@@

>>>>

The multiple imputation method can be roughly divided into 3 steps.
First, each missing value in an incomplete dataset is `imputed' (i.e.,
filled in) several times. Next, the analysis of scientific interest is
performed on each of the completed datasets. And finally, the results of
these analyses are pooled into a single aggregated statistic. The
theoretical foundation of these streps was laid by Rubin
\citeyearpar[1972,][]{rubin87}. For an accessible and comprehensive
introduction to MI from an applied perspective, see e.g.~\citet{buur18}.
In this paper, we focus on the first of these steps: the algorithmic
procedure with which the missing values are imputed.

There is no scientific consensus on the convergence properties of MI
algorithms \citep{taka17}. Some default techniques might not yield
converged states at all \citep{murr18}. Therefore, algorithmic
convergence should be monitored carefully. The current practice is
visually inspecting imputation chains for signs of non-convergence, but
this may be challenging to the untrained eye \citep[\(\S\)
6.5.2]{buur18}. Moreover, visually inspecting imputation chains can only
diagnose severely pathological cases of non-convergence. Therefore,
diagnostic evaluation of convergence would be preferred.

Convergence cannot be determined for sure, because MICE
(\textbf{introduce MICE before!}) is a type of MCMC method, and
convergence of MCMC algorithms is not from a scalar to a point, but from
one distribution to another. Therefore, generated values will vary even
after convergence. The aim of convergence diagnostics for MCMC methods
is thus not to establish the point at which convergence is reached, but
to diagnose severe non-convergence.

Several MCMC convergence diagnostics exist, but it is not known
whetherthese are appropriate for MICE. The application of convergence
diagnostics to MI methods has not been systematically studied
\citep{buur18}. The open questions are, e.g.: How can non-convergence be
diagnosed? How many iterations are sufficient/required? What are the
effects of non-convergence on estimates, predictions and inference? What
is an appropriate non-convergence diagnostic? Are the default number of
iterations sufficient? The defaults of common MI software: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 10
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

@@

The stationarity component of convergence may be evaluated with
autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard
error \citep[or `MC error';][]{gewe92}, and Raftery and Lewis's
\citeyearpar{raft91} procedure to determine the effect of trending
within chains (\textbf{or the chain length required to cancel out
trending?}).

@@

Note that all of these methods evaluate the convergence of univariate
scalar summaries (e.g., chain means or variances). These convergence
diagnostics cannot diagnose convergence of multivariable statistics
(i.e., relations between scalar summaries). \citet{buur18} proposed to
implement multivariable evaluation through eigenvalue decomposition
\citep{mack03}. This method is outside of the scope of the current study
(\textbf{not anymore?}).

@@

\hypertarget{univariate-estimates-and-convergence-diagnostics}{%
\subsection{Univariate estimates and convergence
diagnostics}\label{univariate-estimates-and-convergence-diagnostics}}

The bias in the estimates of the variable means show little to no
difference between simulation conditions. It doesn't seem to matter how
many iterations you use in the mice algorithm, the estimates are
unbiased. Similarly, the bias in the estimated variances is more or less
stable across simulation conditions. These univariate quantities appear
to be unaffected by the number of iterations.

When applied to the imputation chain means, \(\widehat{R}\) indicates
that the mice algorithm does not reach a converged state
(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
the most recent recommended threshold reached (\(\widehat{R}\)
\textless{} 1.01). The conventional \(\widehat{R}\) threshold of 1.2 is
reached in simulation conditions \(T = 3\) and \(T > 6\). (With the
default number of iterations (maxit = 5), this dip in Rhat values would
be spotted, so it is no problem.) The point at which an extra iteration
does not seem to improve the Rhat value is around \(T=30\).

Auto-correlations indicate no sign of trending within imputation chains.
In most simulation conditions AC is smaller than or about equal to zero.
Across simulation conditions, however, the autocorrelation curve does
not trend towards 0. Autocorrelation values plateau off at a value of
around .1 . This is a small positive autocorrelation, which would
indicate some trending within chains.

The \(\widehat{R}\) and \(AC\) values for the imputation chain variances
show equal trends and are therefore not discussed separately. Taken
together, univariate estimates seem robust across simulation conditions.
There is no clear effect of the number of iterations on the bias in
these estimates, while the convergence diagnostics indicate that the
algorithm did not reach a completely converged state (yet).

\hypertarget{multivariate-estimates}{%
\subsection{Multivariate estimates}\label{multivariate-estimates}}

There is a clear bias in the regression estimates in simulation
conditions where the number of iterations is smaller than four. In
simulation conditions where \(T > 5\) there is little to no bias in the
estimated regression coefficients. In most simulation conditions nominal
coverage is obtained (i.e., coverage rates of .95) for the confidence
intervals around the regression coefficients. Conditions with only one
or two iterations show some undercoverage. Since confidence interval
width is stable across conditions, the undercoverage may be attributed
to the bias in the estimated regression coefficients.

If we look at the estimated proportion of explained variance in outcome
variable Y we see that the coefficient of determination is underestimate
estimated in conditions where the number of iterations is equal to two
or less, and slightly overestimated in conditions where the number of
iterations is equal to three or more.

In short, we see that the minimum number of iterations required to
obtain unbiased, confidence valid regression estimates is 5. This value,
however, is dependent on the percentage of missing values. E.g., with
95\% of cases having missing data we need at least seven iterations to
obtain unbiased results.

@@

For 25\% miss, no bias in estimated means, Rh flat from 30 it. (but
\textless1.1 for it\textgreater11), AC never high but stable after 40
it. Bias in estimated variances only if it=1 or maybe 2?, while Rh of
chain variances is flat around it 25 (smaller than 1.1 for
it\textgreater9), and AC never positive, flat around it 30. Bias in
estimated reg. coeff. only pronounced for the first iteration., while
coverage rates are ok immediately on average, variance explained is
marginally underestimated (bias of 0.5\% on an effect of 19.25\%).

\hypertarget{results-old-just-for-reference}{%
\section{Results {[}OLD, JUST FOR
REFERENCE!{]}}\label{results-old-just-for-reference}}
